The `zoon` R package for reproducible and shareable species distribution modelling

### 2016-03-21 ms.Rmd
working on paper structure

### Abstract

@@

### Introduction

@@

Reproducibility crisis in species distribution modelling (SDM).

<!-- from Greg: -->
In 2008, @Beale2008 suggested that “we are currently unable to build useful distribution models for many species”. 
Using null models, @Beale2008 found that chance associations between climate and species’ distributions could account for the modelled relationships which they then suggested undermined the rationale for SDM. 
Amongst the replies to this article was a re-analysis by @Araujo2009 that disputed this finding. 
@Beale2008 put the alternative results to differences in distribution data, climate data and spatial domain, but emphasised the use of different null model distributions. 
Whilst some code was provided in the original publication by @Beale2008, the following papers did not. 
The resolution of this very important discussion was hindered by the lack of data and code sharing.

@@

Core, soluble problems:

@@

<!-- Past work towards this goal & where it falls short - the USP of ZOON and requirements -->

@@

#### The ZOON project

The ZOON project takes a different approach, instead defining only the core architecture of workflows, and interfacing with an open repository of community-contributed software modules.
This paper introduces version 0.4-22 of the `zoon` R package, which encodes the majority of this functionality.
First, we describe the modular structure of zoon workflows, illustrate how to download and upload software modules, and demonstrate how completed `zoon` workflows can be shared, evaluated and extended.
Then we illustrate these concepts, and illustrate the contribution we hope `zoon` will make to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.

<!-- The Solution: ZOOOOOON!  -->
### The `zoon` R package

@@

#### Constructing a `zoon` workflow

@@

```{r, echo = FALSE, fig.cap = "The modular SDM structure encoded by a zoon workflow. A) Flow diagram representing the module types. B) The required inputs and outputs for each module types (full details given in the `zoon` vignette 'Building a module'). C) Chaining and listing modules of the same type. D) An example workflow, corresponding to Example 1 below.\\label{fig:workflows}"}

# set up plotting window
plot.new()
par(mar = rep(0, 4))
plot.window(0:1, 0:1)

@@

# text over the top
text(0.5, 0.5, 'PLACEHOLDER',
     col = grey(0.3),
     cex = 3,
     xpd = NA)
```

Firstly, occurrence data must be read into the software package or downloaded from online repositories.
Occurrence usually takes the form of presence-absence data or presence-only data, though abundance data is also used.
Secondly, predictor variables or covariates (typically environmental covariates) are required and the values of these covariates, at the locations of the occurrence data, must be extracted.
Thirdly, a wide variety of processes are often applied to the occurrence and covariate data.
These processes include data cleaning, data thinning to account for spatial biases, feature selection using PCA or association tests and the splitting of data into training and test sets or cross validation folds.
Fourthly, once the data has been suitable manipulated a model is fitted to estimate the relationships between the covariates and occurrence data.
These models include simple statistical models such as GLMs as well as modern, flexible machine-learning methods such as MaxEnt and boosted regression trees.
Finally, model outputs are needed.
The predictive power and goodness of fit must be assessed and the model parameters or response curves must be examined.
The model is likely to be used to predict species occurrence, either in the vicinity of the occurrence data or elsewhere, or into the past or future.

@@

@@

#### Example 1. A fairly simple SDM, tweaking some different stuff and looking at the results

Explanation & citation

One of the modules we had to write:

```{r, eval=FALSE}
AwesomeModule <- function(.df) {
    something
    return (something else)
}
```

What the workflow looks like

@@

Some tweak we did to it:

@@

While the `zoon` R package can be used in isolation, future work will develop an online platform to facilitate exploring contributed modules and workflows.
This platform will also provide an online space to discuss, and openly evaluate, proposed best practices in SDM.
The zoon R package therefore represents a step towards a more reproducible ecosystem of SDM software.

>>>>

# Abstract

@@

# Introduction

@@

... reproducibility crisis in SDM.

<!-- from Greg, will need trimming down -->
In 2008, @Beale2008 suggested that “we are currently unable to build useful distribution models for many species”. 
Using null models, @Beale2008 found that chance associations between climate and species’ distributions could account for the modelled relationships which they then suggested undermined the rationale for SDM. 
Amongst the replies to this article was a re-analysis by @Araujo2009 that disputed this finding. 
@Beale2008 put the alternative results to differences in distribution data, climate data and spatial domain, but emphasised the use of different null model distributions. 
Whilst some code was provided in the original publication by @Beale2008, the following papers did not. 
The resolution of this very important discussion was hindered by the lack of data and code sharing.

@@

### Solutions to these problems

The core, soluble problems are:

@@

<!-- Past work towards this goal & where it falls short - the USP of ZOON and requirements -->
### Reproducible software projects

@@

### The ZOON project

The ZOON project takes a different approach, instead defining only the core architecture of workflows, and interfacing with an open repository of community-contributed software modules.
This paper introduces version 0.4-22 of the `zoon` R package, which encodes the majority of this functionality.
First, we describe the modular structure of zoon workflows and demonstrate how completed `zoon` workflows can be shared, evaluated and extended.
Then we illustrate these concepts, and the contribution we hope `zoon` will make to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.

<!-- The Solution: ZOOOOOON!  -->
# The `zoon` R package

@@

### Constructing a `zoon` workflow

@@

```{r, echo = FALSE, fig.cap = "The modular SDM structure encoded by a zoon workflow. A) Flow diagram representing the module types. B) The required inputs and outputs for each module types (full details given in the `zoon` vignette 'Building a module'). C) Chaining and listing modules of the same type. D) The structure of a workflow object.\\label{fig:workflows}"}

# four panel plot, no margins
par(mar = c(1, 1, 2, 1),
    mfrow = c(2, 2))

# placeholder plots for each panel
for (i in 1:4) {

  # set up plotting window
  plot.new()
  plot.window(0:1, 0:1)

@@

  # text over the top
  text(0.5, 0.5, 'PLACEHOLDER',
       col = grey(0.3),
       cex = -13,
       xpd = NA)

  # add a panel letter
  mtext(text = LETTERS[i],
        side = 3,
        line = 0.5,
        adj = 0)  
}
```

#### Module types

* *Occurrence* - Usually presence-absence data or presence-only data, though abundance data is also used.
* *Covariates* - Predictor variables or covariates (typically environmental covariates) are required and the values of these covariates, at the locations of the occurrence data, must be extracted.
* *Process* - Processes  applied to the occurrence and covariate data. These processes include data cleaning, data thinning to account for spatial biases, feature selection using PCA or association tests and the splitting of data into training and test sets or cross validation folds.
* *Model* - Once the data has been suitable manipulated a model is fitted to estimate the relationships between the covariates and occurrence data. These models include simple statistical models such as GLMs as well as modern, flexible machine-learning methods such as MaxEnt and boosted regression trees.
* *Output* - The predictive power and goodness of fit must be assessed and the model parameters or response curves must be examined. The model is likely to be used to predict species occurrence, either in the vicinity of the occurrence data or elsewhere, or into the past or future.

@@

### Inspecting, sharing and extending a `zoon` workflow

@@

#### Example 1. Modelling the distribution of ???

??? constructed a species distribution model for ??? using ???.

Such a model can be re-constructed as a `zoon` workflow using modules which are already available in the software repository:

@@

The resulting workflow contains all the c
Some tweak we did to it:

@@

The `zoon` packages provides solutions to some of the technical barriers to the ultimate goal of more open, productive, reproducible SDM research.
Achieving this goal will also depend on overcoming the social hurdle of developing a common tool for comparing and quantitatively evaluating SDMs.

This will require achieving a critical mass of `zoon` modules, so that there is an incentive for species distribution modellers to use the software, and a critical mass of users, so that there is an incentive for SDM methods developers to encode their proposed methods as modules in the community repo.
At this point `zoon` will provide a common environment in which to evaluate SDM methods and enable the SDM research community to make evidence-based decisions about best practice and the goals for development of the field.

In order to facilitate this social aspect, future work will develop an online platform to enable exploring contributed modules and workflows.
This platform will also provide an online space to discuss, and openly evaluate, proposed best practices in SDM.
The zoon R package therefore represents a step towards a more reproducible ecosystem of SDM software.
